GLOBALIZER: A NOVEL SUPERCOMPUTER SOFTWARE SYSTEM FOR SOLVING TIME-CONSUMING GLOBAL OPTIMIZATION PROBLEMS

### 2016-08-04 2016/GOM/paper.tex
Update GOM paper

\author{
\name{V. P. Gergel\textsuperscript{a}$^{\ast}$\thanks{$^\ast$Corresponding author. Email: latex.helpdesk@tandf.co.uk}
and K. A. Barkalov\textsuperscript{b} and A. V. Sysoev}
\affil{\textsuperscript{a}Taylor \& Francis, 4 Park Square, Milton Park, Abingdon, UK;
\textsuperscript{b}Institut f\"{u}r Informatik, Albert-Ludwigs-Universit\"{a}t, Freiburg, Germany}
\received{v5.0 released July 2015}
}

@@

\begin{figure}[h!]
    \centering
		\includegraphics[width=0.55\textwidth]{pictures/peanoCUsage.eps}
		\caption{The computational scheme for obtaining the value of the reduced one-dimensional function \(\varphi(y(x)\)}
    \label{fig:peanoCUsage}
\end{figure}

@@

\subsection{Core Multidimensional Generalized Algorithm of Global Search}
\label{subsec:corepar}
The information-statistical theory of global search formulated in (Strongin, R. G., 1978), (Strongin, R. G., Sergeyev, Y. D., 2000)
has served as a basis for the development of a large number of efficient multiextremal
optimization methods --- see, for example, (Gergel, V. P., 1996, 1997), (Gergel, V. P., Strongin, R. G., 2003, 2005), (Grishagin, V. A., Strongin, R. G., 1984), (Sergeyev Y. D., 1995, 1999), (Sergeyev, Y. D., Grishagin V. A., 1994, 2001), (Sergeyev, Y. D., Strongin, R. G., Lera, D., 2013), (Barkalov K. A., Gergel V. P., 2014), etc.
\par
Multidimensional Algorithm of Global Search (MAGS) established the basis for the
methods applied in Globalizer. The general computational scheme of MAGS can be
presented as follows --- see (Strongin, R. G., 1978), (Strongin, R. G., Sergeyev, Y. D., 2000).
\par
Let us introduce a simpler notation for the problem being solved on a computing node
\begin{equation}
\label{eq:oneDimTask}
f(x) = \varphi(y(x)):x\in [0;1]\}
\end{equation}
\par
The initial iteration of the algorithm is performed at an arbitrary point \(x^1(0,1)\).
Let us assume further \(k\), \(k>1\), iterations of a global search are to be completed.
The selection of the trial point \(k+1\) for the next iteration is performed according to the following rules.
\subsection{Parallel Computations for Systems with Shared Memory}
\label{subsec:sharedpar}

\subsection{Parallel Computations for Systems with Distributed Memory}
\label{subsec:distribpar}

\section{Globalizer System Architecture}
\label{sec:arch}
\section{Results of Numerical Experiments}
\label{sec:experiments}
\subsection{Parallel Computations Using Intel Xeon Phi processors}
\subsection{Parallel Computations Using General Purpose Graphic Processors}

>>>>

\author{
\name{V. P. Gergel\textsuperscript{a}$^{\ast}$\thanks{$^\ast$Corresponding author. Email: gergel@unn.ru}
and K. A. Barkalov\textsuperscript{b} and A. V. Sysoev}
\affil{\textsuperscript{a}Taylor \& Francis, 4 Park Square, Milton Park, Abingdon, UK;
\textsuperscript{b}State University of Nizhny Novgorod, Nizhny Novgorod, Russia}
}

@@

\begin{figure}[h!]
    \centering
		\includegraphics[width=0.55\textwidth]{pictures/peanoCUsage.eps}
		\caption{The computational scheme for obtaining the value of the reduced one-dimensional function \(\varphi(y(x))\)}
    \label{fig:peanoCUsage}
\end{figure}

@@

\subsection{Core Multidimensional Generalized Algorithm of Global Search}
\label{subsec:corepar}
The information-statistical theory of global search formulated in (Strongin, R. G., 1978), (Strongin, R. G., Sergeyev, Y. D., 2000)
has served as a basis for the development of a large number of efficient multiextremal
optimization methods --- see, for example, (Gergel, V. P., 1996, 1997), (Gergel, V. P., Strongin, R. G., 2003, 2005), (Grishagin, V. A., Strongin, R. G., 1984), (Sergeyev Y. D., 1995, 1999), (Sergeyev, Y. D., Grishagin V. A., 1994, 2001), (Sergeyev, Y. D., Strongin, R. G., Lera, D., 2013), (Barkalov K. A., Gergel V. P., 2014), etc.
Multidimensional Algorithm of Global Search (MAGS) established the basis for the
methods applied in Globalizer. The general computational scheme of MAGS can be
presented as follows --- see (Strongin, R. G., 1978), (Strongin, R. G., Sergeyev, Y. D., 2000).
Let us introduce a simpler notation for the problem being solved on a computing node
\label{eq:oneDimFunc}
f(x) = \varphi(y(x)):x\in [0,1]\}
The initial iteration of the algorithm is performed at an arbitrary point \(x^1(0,1)\).
Let us assume further \(k\), \(k>1\), iterations of a global search are to be completed.
The selection of the trial point \(k+1\) for the next iteration is performed according to the following rules.
\textit{Rule 1}. To renumber the points for the preceding trials by the lower indices in order of increasing value of coordinates
  \label{step1}
0=x_0<x_1<\dotsc<x_{k+1}=1
the points \(x_0\), \(x_k\) were introduced additionally for convenience in further explanation,
the values of the minimized function \(z_0\), \(z_k\) at these points are undefined.
\textit{Rule 2}. To compute a current estimate of the Hölder constant \(H\) from (\ref{eq:holder})
  \label{step2}
m=\max_{1\leqslant i\leqslant k}\dfrac{|z_i-z_{i-1}|}{\rho_i},
\begin{matrix}
    M =
    \left\{
    \begin{matrix}
    r\cdot u,m>0 \\
    1,m=0
    \end{matrix} \right.
    \end{matrix}
as the maximum of the relative differences of the minimized function \(f(x)\) from (\ref{eq:oneDimFunc})
on the set of executed trial points \(x_i,1\leqslant i\leqslant k\) from (\ref{step1}). Further, \(\rho_i=(x_i-x_{i-1})^\frac{1}{N},1\leqslant i\leqslant k+1\).
The constant \(r\), \(r>1\), is the \textit{parameter} for the algorithm.
\textit{Rule 3.} For each interval \((x_{i-1},x_i),1\leqslant i\leqslant k+1\), compute the characteristics \(R(i)\) where
\label{step3}
  R(i)=\rho_i+\dfrac{(z_i-z_{i-1})^2}{M^2\rho_i}-2\dfrac{z_i+z_{i-1}}{M},\quad 1<i<k+1 \\
  R(i)=2\rho_1-4\dfrac{z_1}{M},\quad i=1 \\
  R(i)=2\rho_{k+1}-4\dfrac{z_k}{M},\quad i=k+1
\textit{Rule 4.} To determine the interval with the maximum characteristic
\label{step4}
R(t)=\max_{1\leqslant i \leqslant k+1}R(i)
\textit{Rule 5.} To execute a new trial (computation of the minimized function value \(f(x)\))
at the point \(x^{k+1}\) located within the interval with the maximum characteristic from (\ref{step4})
\label{step5}
  x_{k+1}=\dfrac{x_t+x_{t-1}}{2}-sign(z_{t}-z_{t-1})\dfrac{1}{2r}\left[\dfrac{|z_{t}-z_{t-1}|}{\mu}\right]^N,\quad 1<t<k+1 \\
  x_{k+1}=\dfrac{x_t+x_{t-1}}{2},\quad t=k+1
The stop condition by which the trials are terminated, is defined by the inequality
  \label{eq:stop}
\rho_t<\varepsilon
for the interval with the maximum characteristic from (\ref{step4}) and \(\varepsilon >0\) is the predefined
accuracy of the problem solution. If the stop condition is not satisfied, the index \(k\) is incremented by unity,
and a new global search iteration is executed.
In order to explain the algorithm presented above, let us note the following.
The characteristics \(R(i), 1\leqslant i\leqslant k+1\), calculated according to (\ref{step3})
could be interpreted as some measures of importance to the intervals with respect to
location of the global minimum point. Thus, the scheme for selecting the interval
for executing the next trial described in (\ref{step4}-\ref{step5}) becomes more clear --- the point of every next
global search iteration is selected within the interval where the global minimum point can most likely be found.
The conditions for the algorithm’s convergence described above were examined, for example, in (Strongin, R. G., Sergeyev, Y. D., 2000).

\subsection{Parallel Computations for Systems with Shared Memory}
\label{subsec:sharedpar}
Modern supercomputer computational systems consist of plenty of computer nodes,
which include several multicore processors. At that, random access memory is
shared for the computing nodes --- the content of any memory element can be read (written)
by any computer cores available at any arbitrary moment. In most cases, shared memory is
uniform --- the time characteristics for accessing memory are the same for all computer
cores and for all memory elements.
\par
The following speculations can form the basis for selecting parallel computation
organization methods. As was mentioned above when describing the MAGS algorithm,
the characteristics \(R(i),1\leqslant i\leqslant k+1\), calculated according to (\ref{step3})
can be interpreted as some measures of the importance of intervals with respect to the location of
the global minimum point. Following this understanding, each successive global search
iteration is executed within the most important interval with the maximum value of the
characteristic. So far, in this case it is obvious how to select the other intervals for
organizing simultaneous computations of the minimized function values at several different
points within the search domain as well --- these should be the intervals with the next
magnitudes of characteristics.
\par
The computational scheme of the Parallel Multidimensional Algorithm of Global Search for
computer systems with shared memory (PMAGS) was developed based the approach considered
above and is practically identical to the MAGS scheme --- the differences consist just in the
following set of rules --- see (Strongin, R. G., Sergeyev, Y. D., 2000), (Strongin, R. G., K. A., Gergel, V. P., Grishagin, V. A., Barkalov, K. A.,  2013).
\par
\textit{Rule 4'}. To arrange the characteristics of the intervals obtained according to (\ref{step3}) in decreasing order
\begin{equation}
\label{step4par}
R(t_1)\geqslant R(t_2)\geqslant \dots \geqslant R(t_{k})\geqslant R(t_{k+1})
\end{equation}
and to select \(p\) intervals with the indices \(t_j,1\leqslant j\leqslant p\), with
the highest values of characteristics (\(p\) is the number of processors/cores used for the parallel computations).
\par
\textit{Rule 5'}. To execute new trials (computations of values for the minimized
function \(f(x)\)) at the points \(x_{k+j},1\leqslant j\leqslant p\), located in the
intervals with the highest characteristics from (\ref{step4par})
\begin{equation}
\label{step5par}
\begin{array}{cr}\\
x_{k+j}=\dfrac{x_{t_j}+x_{t_j-1}}{2}-sign(z_{t_j}-z_{t_j-1})\dfrac{1}{2r}\left[\dfrac{|z_{t_j}-z_{t_j-1}|}{\mu}\right]^N,\quad 1<t_j<k+1 \\
x_{k+j}=\dfrac{x_{t_j}+x_{t_j-1}}{2},\quad t_j=1,t_j=k+1
\end{array}
\end{equation}
\par
The stop condition for the algorithm (\ref{eq:stop}) should be checked for all
intervals from (\ref{step4par}), in which the scheduled trials are executed
\begin{equation}
  \label{eq:stop}
\rho_{t_j}<\varepsilon,1\leqslant j\leqslant p
\end{equation}
As before, if the stop condition is not fulfilled, the index \(k\) is
incremented by \(p\), and a new global search iteration is executed.
\par
The conditions for the developed parallel algorithm’s convergence and for the
non-redundancy of parallel computations have been examined in (Strongin, R. G., Sergeyev, Y. D., 2000).
Thus, in particular, when the conditions for convergence are satisfied, the parallel
computations are non-redundant as compared to the sequential method when using up to
\(2N\) processors/cores (where \(N\) is the dimensionality of the global optimization problem being solved).

\subsection{Parallel Computations for Systems with Distributed Memory}
\label{subsec:distribpar}
The next level of parallel computations in high-performance computer systems consists of
using several computational nodes. For that, each computational node has its own separate
memory and, in this case, the interaction between different nodes can be provided by
means of data transfer only via the computer communication network.
To organize parallel computations for multiprocessor systems with distributed memory, using a set of mappings
  \label{eq:maps}
Y_s(x)=\{y^1(x),\dots,y^s(x)\}
instead of applying a single Peano curve \(y(x)\) has been proposed in (Strongin, R. G., 1992, Strongin, R. G., Sergeyev, Y. D., 2000).
In order to build the set \(Y_s(x)\), several different approaches can be applied.
Thus, for example, in (Strongin, R. G., 1992) a scheme has been applied by which each
mapping \(y_i(x)\) from \(Y_s(x)\) is built as the result of some shift along the main
diagonal of the hypercube \(D\). This way, the set of constructed Peano curves enables the
close prototypes \(x'\), \(x''\)  to be obtained from the interval \([0,1]\) for any close
multidimensional images \(y'\), \(y''\) from \(D\) differing in one coordinate,
only for some mapping \(y_k(x), 1\leqslant k\leqslant s\).
Some other methods for constructing multiple mappings have been examined in (Strongin, R. G., K. A., Gergel, V. P., Grishagin, V. A., Barkalov, K. A.,  2013).
The set of mappings \(Y_s(x)\) from (\ref{eq:maps}), for multidimensional problem (\ref{eq:task})
generates \(s\) subsidiary information-linked one-dimensional problems of the kind (\ref{eq:oneDimTask}):
  \label{eq:oneDimProblemSerie}
  \varphi(y^l(x^l))=\min\{\varphi(y^l(x)):x\in [0,1]\},1\leqslant l\leqslant s
It is important to note that the family of one-dimensional problems \(\varphi(y^l(x)),1\leqslant l\leqslant s\),
obtained as a result of the dimensional reduction is information-linked --- the function
values computed for any problem \(\varphi(y^l(x))\) from the family (\ref{eq:oneDimProblemSerie}) can be used for all
of the remaining problems of this family.
The information compatibility of the problems from the family (\ref{eq:oneDimProblemSerie}) allows the parallel
computations to be organized in an obvious enough way. Thus, each particular problem can
be solved by a separate processor in the computing system; the exchange of search information
between the processors should be performed during the course of the computations.
As a result of using such a parallelization method, a unified approach for organizing parallel
computations for a multiprocessor computer systems with distributed, as well as shared
memory can be developed. This method consists of the following.
\begin{enumerate}
  \item The family of reduced one-dimensional information-linked problems (\ref{eq:oneDimProblemSerie}) is
  distributed among the computational nodes of a multiprocessor system. One problem,
  as well as several others from within the family, can be allocated to each particular computational node.
  \item The Parallel Multidimensional Algorithm of Global Search (PMAGS, see Subsection \ref{subsec:sharedpar})
  is applied to solve the allocated problems from the family (\ref{eq:oneDimProblemSerie})
  at each computational node, supplemented by the following rules of information interaction.
  \begin{enumerate}
    \item Prior to beginning a new trial at any point \(x'\in [0,1]\) for any problem \(\varphi(y_l(x)),1\leqslant l\leqslant s\),
    the following should be performed:
      \item compute the image \(y'\in D\) for the point \(x'\in [0, 1]\) according to the mapping \(y_l(x)\),
      \item determine  the prototypes \(x_i',1\leqslant i\leqslant s\), for each
      problem from the family (\ref{eq:oneDimProblemSerie}),
      \item send the prototypes \(x_i',1\leqslant i\leqslant s\) to all computational
      nodes employed in order to exclude the repeated selection of the intervals in
      which the prototypes fall, for using them to determine the points for new trials.
      To organize the data transfer, a queue can be formed at each computational
      node for sending the trial points and receiving the minimized function values at these points.
    \item After completing any trial for any problem \(\varphi(y_l(x)),1\leqslant l\leqslant s\), at any point
    \(x'\in[0,1]\), it is necessary:
      \item to determine all prototypes \(x_i',1\leqslant i\leqslant s\), for the point of the completed trial
      for each problem from the family (\ref{eq:oneDimProblemSerie}),
      \item to send the prototypes \(x_i',1\leqslant i\leqslant s\), and the result of the
      trial \(z'=\varphi(y_l(x'))\) to all computational nodes employed to include the
      obtained data into the search information processed within the rules of the parallel global search algorithm.
    \item Prior to starting the next global search iteration, the algorithm should
    check the queue of the received data; if there are some data in the queue,
    they should be included in the search information.
\end{enumerate}
\end{enumerate}
The possibility of asynchronous data transfer (the computational nodes process received
data only upon receipt) is a meaningful feature of such a scheme for organizing parallel computations.
In addition, there is no single managing node within this scheme. The number of
computational nodes can vary during the course of the global search, and excluding any
node does not result in the loss of the sought global minimum of the minimized multiextremal function.
This computational scheme generates the Generalized Parallel Multidimensional Algorithm of
Global Search (GPMAGS) for high-performance computing systems, which may include many
multiprocessor (multicore) computational nodes with distributed memory as well as
accelerators for computations based on Intel Xeon Phi processors and on general purpose graphic processors.
Additional information on organizing such parallel computation schemes is provided in (Gergel, V. P., Sidorov, S., 2015).

\section{Globalizer System Architecture}
\label{sec:arch}
As previously mentioned in Section \ref{sec:intro}, the development of software systems
for global optimization is complex professional work. The optimization systems should
include advanced optimization methods, provide for the accumulation and efficient
processing of search information, include various forms of visualization for the results,
provide tools for interacting with users, etc. All the features pointed out above
become much more complicated when organizing parallel computations using high-performance
multiprocessor systems of varying architectures.
\par
The Globalizer considered in this paper expands the family of global optimization
software systems successively developed by the authors during the past several years.
One of the first developments was the SYMOP multiextremal optimization system (see Gergel, V. P., 1993),
which has been successfully applied for solving many optimization problems. A special
place is occupied by the ExaMin system (see, for example, (Barkalov, K., Gergel V., 2015)),
which was developed and used extensively to investigate the application of novel parallel
algorithms to solve global optimization problems using high-performance multiprocessor computing systems.

\begin{figure}[h!]
    \centering
		\includegraphics[width=0.75\textwidth]{pictures/globalizerScheme.eps}
		\caption{The Globalizer system architecture (Blocks 1-2, 5-7 have been implemented; Blocks 3-4 and 8-11 are under development)}
    \label{fig:globalizerScheme}
\end{figure}
The Globalizer system architecture is presented in Figure \ref{fig:globalizerScheme}. The structural components are:
  \item Block 0 consists of the procedures for computing the function values (criteria
  and constraints) for the optimization problem being solved; this is an external
  block with respect to the Globalizer system and is provided by the user per a predefined interface.
  \item Blocks 1-4 form the optimization subsystem and solve the global optimization
  problems (Block 1), nonlinear programming (Block 2), multicriterial optimization (Block 3),
  and general decision making problems (Block 4). It is worth noting the sequential
  scheme of interaction between these components --- the decision making problems are
  solved using the multicriterial optimization block, which, in turn, uses the nonlinear programming block, etc.
  \item Block 5 is a subsystem for accumulating and processing the search information;
  this is one of the basic subsystems --- the amount of search information for complex
  optimization problems may appear to be quite large on the one hand, but, on the other
  hand, the efficiency of the global optimization methods depends to a great extent on how
  completely all of the available search data is utilized.
  \item Block 6 contains the dimensional reduction procedures based on Peano evolvents;
  the optimization blocks solve the reduced (one-dimensional) optimization problems;
  Block 6 provides interaction between the optimization blocks and the initial
  multidimensional optimization problem.
  \item Block 7 organizes the choice of parallel computation schemes in the Globalizer
  system subject to the computing system architecture employed (the numbers of cores in the
  processors, the availability of shared and distributed memory, the availability of
  accelerators for computations, etc.) and the global optimization methods applied.
  \item Block 8 is responsible for managing the parallel processes when performing the
  global search (determining the optimal configuration of parallel processes, distributing
  the processes between computer elements, balancing computational loads, etc.).
  \item Block 9 is a common managing subsystem, which fully controls the whole
  computational process when solving global optimization problems.
  \item Block 10 is responsible for organizing the dialog interaction with users for
  stating the optimization problem, adjusting system parameters (if necessary), and
  visualizing and presenting the global search results.
  \item Block 11 is a set of tools for visualizing and presenting the global search results;
  the availability of tools for visually presenting the computational results enables
  the user to provide efficient control over the global optimization process.
The Globalizer system comes in two different options:
  \item Globalizer Pro is a research version of the system providing a full set of
  capabilities that are oriented toward solving of the most complicated global optimization problems.
  \item Globalizer Lite is a limited version of the Globalizer system accessible for free for
  solving global optimization problems with a medium level of difficulty, and is accessible free of charge.
\section{Results of Numerical Experiments}
\label{sec:experiments}
In order to confirm the operability of the Globalizer system, results from two
series of computational experiments are presented in this section.
The computational experiments were conducted using the Lobachevsky supercomputer at the
State University of Nizhny Novgorod (operation system --- CentOS 6.4, management system --- SLURM).
Each supercomputer node had 2 Intel Sandy Bridge E5-2660 processors, 2.2 GHz, with 64 GB RAM.
Each processor had 8 cores (i.e. a total of 16 CPU cores were available at each node).
In order to generate the executable program code, the Intel C++ 14.0.2 compiler was used.
A number of nodes in the computer system utilized Intel Xeon Phi 5110P processors,
each containing 60 cores (a total of 240 threads). In addition, the supercomputer also
included computer nodes equipped with two GPU NVIDIA Kepler K20Х, each providing 14
stream multiprocessors (a total of 2688 CUDA-cores). In order to generate the executable
code, the CUDA Toolkit 6.0 was used.
The problems generated by the GKLS-generator (Gaviano, M., Lera, D., Kvasov, D. E., Sergeyev, Y. D., 2003)
were selected for the test problems. This generator allowed for obtaining the multiextremal
optimization problems with \textit{a priori} known properties: the number of local minima,
the sizes of their attractors, the global minimum point, the function value at this point, etc.
In order to simulate the computation costs inherent in the applied optimization problems
(see, for example, Bastrakov, S., Gergel, V., Meyerov, I., et al, 2013), in all of the experiments
computing the objective function was complicated by auxiliary computations, not altering the
form of the function and the position of its minima.
We considered the global minimum \(y^*\) in the test problem to be found if the
algorithm generated a trial point \(y^k\) in the \(\delta\)-nearness of the global
minimum i. e. \(\Vert y^k-y^* \Vert \leqslant \delta\). The size of the nearness was
selected as \(\delta=\Vert b-a \Vert\sqrt[N]{\Delta}\) where \(a\) and \(b\) are
boundaries of the search domain \(D\), \(N\) is the dimensionality of the optimization problem.
The maximum allowed number of parallel iterations was \(K_{max} = 10000\), and the evolvent
density parameter was \(m=10\).The computational experiments considered below were first
carried out using the ExaMin system (see (Barkalov, K., Gergel V., 2015), (Barkalov, K., Gergel, V., Lebedev, I., 2015), (Barkalov, K., Gergel, V., Lebedev, I., Sysoev, A., 2015), (Gergel, V., Lebedev, I., 2015))
and then were reproduced using the Globalizer system.

\subsection{Parallel Computations Using Intel Xeon Phi processors}
The results of the numerical experiments with GPMAGS on an Intel Xeon Phi are provided in Table 1.
The computations were performed using the Simple and Hard function classes with the dimensions equal to 4 and 5. The parameters  at  and  at  were used. When using the GPMAGS method, the parameter  was selected for the Simple class, and  was selected for the Hard class.
\par
In the first series of experiments, serial computations using MAGS were executed. The average number of iterations performed by the method for solving a series of problems for each of these classes is shown in row I. The symbol “>” reflects the situation where not all problems of a given class were solved by a given method. It means that the algorithm was stopped once the maximum allowable number of iterations Kmax was achieved. In this case, the Kmax value was used for calculating the average number of iterations corresponding to the lower estimate of this average value. The number of unsolved problems is specified in brackets.
\par
In the second series of experiments, parallel computations were executed on a CPU. The relative “speedup” in iterations achieved is shown in row II; the speedup of parallel computations was measured in relation to the sequential computations (p=1).

\subsection{Parallel Computations Using General Purpose Graphic Processors}
